|

### 2025-01-12 paper/manuscript.Rmd
TK edits.

---
title: |
  | The Promises and Pitfalls of Using Panel Data to Understand Individual Belief Change[^thanks]
author:
  - Turgut Keskintürk[^TK]
  - Pablo Bello[^PB]
  - Stephen Vaisey[^SV]
date: "`r format(Sys.time(), '%B %Y')`"
output: pdf_document
bibliography: ./helpers/references.bib
csl: ./helpers/asa.csl
abstract: "We investigate whether studies on political belief change can identify change trajectories at the individual level. Using simulations and case studies, we propose a grid-search framework that allows researchers to evaluate the extent to which their target estimates generalize to individuals in their study population. We use simulated datasets to estimate plausible values for how many people changed, how much they changed, and who changed, based on observed survey response trajectories. Our results suggest that researchers should think carefully about the conditions under which they may make claims about belief change at the individual level. We provide an R package to help researchers implement this procedure in their own work."
fontsize: 11pt
fontfamily: mathpazo
geometry: margin=1in
figsintext: yes
header-includes:
- \usepackage{array}
- \usepackage{float}
- \usepackage{caption}
- \usepackage{amsmath}
- \captionsetup{justification=raggedright,
                singlelinecheck=false}
---

@@

This *group-to-person generalizability problem*, the claim that group-level findings may not generalize to each person [@mcmanus2023], has important implications for what we can and can't say about political change across the life-course. While the previous research quantified change with various methods---ranging from assessing the time order of longitudinal observations [@kiley2020; @vaisey2020] to multilevel decomposition of over-time variance [@lersch2023]---these studies have been unable to disentangle whether they find a small amount of change in a large number of people, or a large amount of change in a small number of people. In other words, while social science researchers have become sophisticated in evaluating the average levels of change over time in a population, they have mostly overlooked the fact that these average changes can be produced by different processes.

@@

While the cross-sectional processes have received strong scholarly attention in the literature [@vaisey2016; @restrepoochoa2024; @bartels2014], questions of the second sort remained largely unexplored. However, given that theories of political ideology involve theoretical expectations about individual-level outcomes, the fact that we lack information about whether our models capture person-level empirical processes is detrimental to understanding adult political belief change. Therefore, our article aims to address this gap by examining whether we can identify (a) *how many people changed*, (b) *how much they changed*, and (c) *who changed*, based on observed survey response trajectories of political beliefs in empirical data.

Using simulations and empirical case studies, we show that it is indeed possible to use panel data to provide plausible estimates of which generative models are more likely to have produced particular empirical observations. We also find, however, that a reliable identification of *who* changed occurs only under very narrow conditions. In this sense, the modeling individual change problem is very much like the age-period-cohort identification problem that arises in repeated cross-sectional data [CITATION TO FOSSE]. In both cases, the problem cannot be "solved" by any particular methodological trick. Thus, our objective is not to provide a methodological panacea to eliminate the problem, but rather to provide a concrete approach that can help us understand the issue more clearly and manage it more effectively.

This approach builds on the "Approximate Bayesian Computation" (ABC) literature [@beaumont2010] and provides a simple grid-search procedure to help researchers evaluate the extent to which their target estimates might be generalized to a study population. Using this procedure, we show that it is possible to approximate to the underlying data generation processes for *rate of change*, *strength of change*, and *direction of change* in a panel dataset. We provide an R package, `gridsearch`, to help researchers implement this procedure in their own work.

We emphasize the existence of striking heterogeneity in these underlying processes even for a seemingly straightforward case like the marijuana legalization in the GSS; that is, we show that theoretically opposite claims---e.g., large swaths of the population increasing their support for the policy, as opposed to a small group of people completely switching their positions---may very well account for the same data. This finding is particularly important in the longitudinal panel context, which, for most applications, represents the gold standard in evaluating change in observational research. For this reason, we argue that when the meaning of "political change" is not clearly specified, our current empirical strategies could potentially mislead rather than inform, especially if we overlook the underlying generative mechanisms.

We begin with a general overview of the "latent variable model" of beliefs, the problems of reliability and resolution in survey measurement, and the compositional sources of aggregate change. We then introduce our grid-search procedure in the context of three-wave panel data with binary outcomes. Next, we relax this assumption and investigate how linear trajectories at the individual level can assist us in contexts with more than three waves. In the final section, we analyze whether this procedure helps us in classification tasks, particularly in evaluating *who* changed. We conclude by discussing the promises and pitfalls of panel data for understanding individual political change.

# Measuring Change in Political Beliefs

@@

The latent-variable approach acknowledges a fundamental problem in survey research: one's true score $Y_i$ is almost always realized with a certain degree of measurement error. Call this realization the *observed score*, $y_i$. We often assume that $y_i$ represents the expected value of hypothetical draws from $Y_i$, with an $\epsilon$ term reflecting the measurement error in this process---i.e., $y_i = Y_i + \epsilon_i$. This error might result from various factors that introduce biases in how we extract information from individuals, given that problems like question construction and interviewer effects might influence how individuals reveal their true preferences [@alwin1991]. The inherent randomness in the measurement process implies that each draw from $Y_i$ might at least slightly differ from the true score, and this error term is difficult to circumvent with a one-item measure, which is often the case in large public opinion surveys.[^scales] In this sense, there is often a degree of systematic gap between $Y_i$ and $y_i$.

[^scales]: The idea that we should have repeated measures of $Y_i$ is the main intuition behind scale construction [@ansolabehere2008], with the explicit expectation that multiple measures can help us reduce measurement error.

@@

This quantity is a natural candidate for describing aggregate belief change, but, under the hood, it summarizes three compositional processes in one metric. Let us define these processes intuitively. The first component is $R$, or the *rate of change*. This represents the proportion of individuals in our dataset with $\delta y_{i, t_0 \to t_1} \neq 0$, i.e., the proportion of individuals who changed in this period. The second component is $S$, the *strength of change*. This represents the "average" magnitude of change among those who changed.[^heterogeneity] The third component is $D$, the *direction of change*. This represents the net directionality of change, i.e., the difference in proportions of positive and negative changers in our observations. If we combine these three components, we can represent $\bar{y}$ simply as

@@

[^heterogeneity]: Of course, we might further decompose this average strength to a distribution of heterogeneous change trajectories across individuals. For analytical and expository clarity, we bracket this issue from in this article.

@@

As if this weren't complicated enough, we must complicate the issue further by, once again, recognizing the measurement problem of mapping true scores to observed scores. This process often fails in at least two ways: (1) our observed measures might have low *reliability*, such that a hypothetical set of repeated draws may significantly differ from one's true score, and (2) question response categories may be imperfect: the *resolution* of response categories may fail to distinguish consequential differences in true scores. Therefore, the average change captured in $\bar{y}$ may represent a biased or imperfect estimate of change to begin with.

@@

## A Grid-Search Procedure

We build on the Approximate Bayesian Computation (ABC) literature [@beaumont2010] and propose a grid-search procedure to answer this research question. The ABC approach involves simulating parameter values given the data and evaluating the distribution of these parameters that potentially generated our observations. We build a rejection sampler algorithm that

1) Simulates data with several parameters of interest for our DGPs---rate of change, strength of change, direction of change, and reliability of response measurement using

@@

2) Calculates the distribution of response trajectories in both the observed and simulated data, by either tabulating response patterns across time or estimating slope coefficients;

3) Compares the distribution of simulated trajectories to the distribution of observed trajectories, calculating a "distance" score to quantify the extent to which specific simulated values reproduce the observed values;

@@

4) Rejects the simulated data if the distance measure is above a certain pre-specified "caliper" value, and accepts it otherwise;

5) Replicates this analysis $N$ times by drawing samples from the specified priors;

6) Inspects the parameter combinations that most closely approximate to the observed values.

@@

[^parameters]: We define additional parameters that we fix in the case studies presented below: $N$ refers to the number of actors in the fake-data simulation, $T$ refers to the number of waves we have in the panel study, *balance* refers to the observed marginals in the real datasets, allowing us to "cut" the continuous distribution to 0s and 1s, and *resolution* refers to the number of categories to simplify the latent variable, which is restricted to the binary case in this article.

In the next four sections, we apply this procedure in different contexts. In all cases, we restrict our attention to latent variables with binary outcomes to simplify our analyses. The replication files for the article are stored in https://osf.io/AAAAA. The R package for the use of this approach, `gridsearch`, can be accessed via the replication repo or https://tkeskinturk.github.io/gridsearch/.

\newpage

# Study 1: Three-Wave Panel Context

In the three-wave panel context with binary outcomes, the response trajectories are easy to inspect, given that there are only $2^3 = 8$ possible response patterns for each person:

$$
\bigg( \{0,0,0\}, \{0,0,1\}, \{0,1,0\}, \{0,1,1\}, \{1,0,0\}, \{1,0,1\}, \{1,1,0\}, \{1,1,1\} \bigg)
$$

To examine the distribution of these trajectories in the observed data, we can count the number of respondents who go through trajectory $j$. For instance, the marijuana legalization example from Figure 1 has the following contingency table of counts:

$$
\begin{bmatrix}
0, 0, 0 \\
0, 0, 1 \\
0, 1, 0 \\
0, 1, 1 \\
1, 0, 0 \\
1, 0, 1 \\
1, 1, 0 \\
1, 1, 1
\end{bmatrix} 
=
\begin{bmatrix}
869 \\
153 \\
75 \\
115 \\
65 \\
71 \\
54 \\
746 \\
\end{bmatrix}
$$

Approximately 75% of the respondents remain stable, following trajectories of $\{0,0,0\}$ and $\{1,1,1\}$, while others appear to *change* in some respect. The question is the extent to which this contingency table can be approximated by an underlying generative model, where we know the true change parameters.

Let us provide one such example. We simulate a panel dataset, with 20% of the respondents changing their beliefs on marijuana legalization, operationalized as a latent construct with $\mu = 0$ and $\sigma = 1$, by 2 standard deviations in the "negative" direction---note that this is the opposite of the mean pattern observed in Figure 1. We fix all the parameters in the DGP to their observed values: 2,148 individuals observed over 3 periods, with a reliability score of 0.91 [see @hout2016 for the calculation of this value]. We know that the marginals were equal to 0.47 in the entire panel, so we constrain the binary outcomes such that the final distribution of $y$ equals to 0.47 as well. Here are the generated counts for all trajectories, with differences from the original data appended:

$$
\begin{bmatrix}
\\
0, 0, 0 \\
0, 0, 1 \\
0, 1, 0 \\
0, 1, 1 \\
1, 0, 0 \\
1, 0, 1 \\
1, 1, 0 \\
1, 1, 1
\end{bmatrix} 
=
\begin{bmatrix}
\text{Observed} \\
869 \\
153 \\
75 \\
115 \\
65 \\
71 \\
54 \\
746 \\
\end{bmatrix}
-
\begin{bmatrix}
\text{Simulated} \\
810 \\
75 \\
103 \\
57 \\
188 \\
66 \\
148 \\
701 \\
\end{bmatrix}
=
\begin{bmatrix}
\text{Absolute Difference} \\
59 \\
78 \\
28 \\
58 \\
123 \\
5 \\
94 \\
45 \\
\end{bmatrix}
$$

The generative model does a pretty poor job of reproducing the contingency table. 23% of the respondents ($N = 490$) are assigned to the wrong trajectory. This means that we should be able to improve our predictions significantly if we switch to the true data generation model. Unfortunately, we don't know it; the set of true change parameters is the exact thing we want to derive using observed trajectories.

How can we improve our prediction accuracy? Our grid-search procedure allows us to estimate plausible approximation values across a grid of potential generative models. Let us now apply this procedure to the case of marijuana legalization to see whether we can retrieve plausible DGPs.

## The Case of Marijuana Legalization

We first define 2,100 potential generative models using a combination of (1) rate of change, where we vary values from 0 to 1 in increments of 0.05, (2) strength of change, where we vary values from 0 to 2 in increments of 0.1, and (3) five directional possibilities---all negative, all positive, half negative and half positive, 75% negative and 25% positive, and 75% positive and 25% negative. This results in $21 \times 20 \times 5 =$ 2,100 cells, each representing one potential data-generation process that might have produced the observed data. We then generate corresponding panel datasets based on each model. Due to the stochastic nature of our fake-data simulations, we repeat this procedure 1,000 times for each cell, resulting in 2,100,000 datasets for which we evaluate the observed trajectories.

Once we have our fake-data simulations, we count the number of respondents assigned to each response trajectory, calculate the absolute errors relative to the observed data, and normalize these counts by dividing the error by the total number of respondents. We then average these percent errors across 1,000 simulations to obtain stable estimates. Figure 2 presents the results.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure2, out.width = "100%"}
knitr::include_graphics("../figures/grass02.png")
```
```{=latex}
\end{center}
\caption*{Figure 2: Fake-Data Simulations for Marijuana Legalization}
\footnotesize{\textit{Notes:} The figure depicts the extent to which specific parameter combinations approximate the observed data. Each point represents our distance measure, calculated as the sum of absolute errors between observed and simulated contingency tables, where darker colors represent smaller distances. We used 5\%, 10\%, and 20\% distance cuts for color shades, along with a second filter using quantiles of 0.05, 0.10, and 0.20 at each level. The shading thus reflects both absolute and relative accuracy among the simulations. The single yellow point represents the parameter combination that best approximates the observed data in the General Social Survey.}
\end{figure}
```

There are several notable findings. First, error values range from 4.6% to 69.6%, with an interquartile range of 10.6% and 24%. This indicates that many cases fail to achieve a reasonable approximation, but we still end up with reliably close generative models. Second, the best value occurs when the rate of change equals to 25%, with a positive strength of 0.9 standard deviation change (SD). That said, note the black values clustered around the yellow cell in the bottom right panel, where cells range from 25% to 35% rate of change, with strength ranging, once again, from 0.7 to 1 SD. These patterns suggest that the most plausible data generation process occurs when roughly 2 to 4 out of 10 people positively change their approvals by roughly 1 SD. Finally, as expected, there are no plausible estimates where change occurs in the negative direction, providing a plausibility check for our proposed method.

These findings demonstrate the utility of our approach. By iterating over DGPs, we can identify plausible values that may have produced the observed change trajectories in the panel context.

## The Case of the Legality of Abortion

What about cases where the direction or even existence of change is less clear-cut? We now turn our attention to the legality of abortion debate, a contentious issue characterized by high disagreement and high political sorting in the American context [@baldassarri2008; @hout2022]. In 2006, 2008, and 2010 GSS panels, respondents agreed or disagreed with the statement: "it should be possible for a pregnant woman to obtain a legal abortion if the woman wants it for any reason." Figure 3 shows response trajectories for this question across three panels, indicating a clear stasis in the marginals.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure3, out.width = "100%"}
knitr::include_graphics("../figures/abany01.png")
```
```{=latex}
\end{center}
\caption*{Figure 3: The Percent Support for Abortion}
\footnotesize{\textit{Notes:} The figure depicts the percent of American adults saying "it should be possible for a pregnant woman to obtain a legal abortion if the woman wants it for any reason" in the General Social Survey 2006, 2008, and 2010 panel studies. We included respondents who participated in all three waves of each panel study and weighted the results using GSS post-stratification weights. Note that panel attrition caused compositional changes in the estimates, and compared to the GSS time-series data, the estimates of support are marginally inflated.}
\end{figure}
```

However, the contingency table for trajectories shows more heterogeneity. Again, there is considerable stability---approximately 73% of respondents remain stable, following trajectories of $\{0,0,0\}$ and $\{1,1,1\}$, similar to the case of marijuana legalization. But there is more homogeneity in terms of directional change: while trajectories like $\{0,0,1\}$ and $\{0,1,1\}$, cases indicating an upward trend, were 2.27 times more likely in the marijuana legalization case, we now observe similar numbers in the abortion case---230 clear upward trends and 209 clear downward trends. This suggests that the directional heterogeneity of political change might have cancelled the trends in the aggregate, resulting in stable marginals in the observed frequencies.

$$
\begin{bmatrix}
0, 0, 0 \\
0, 0, 1 \\
0, 1, 0 \\
0, 1, 1 \\
1, 0, 0 \\
1, 0, 1 \\
1, 1, 0 \\
1, 1, 1
\end{bmatrix} 
=
\begin{bmatrix}
972 \\
115 \\
101 \\
115 \\
118 \\
99 \\
91 \\
733 \\
\end{bmatrix}
$$

Figure 4 presents the results from our grid-search exercise, implemented across 2,100,000 datasets with 2,100 generative models. We set the reliability to .86 following @hout2016. This time, the estimates are much closer to the observed frequencies, with error values ranging from 3% to 64.5%, and an interquartile range of 5.1% and 23.5%. That said, there is strong variability in the plausible generative scenarios. Note how, in the top right corner, there is a plausibility bound ranging from a rate of change as low as 10% to as high as 100%, and strength of change ranging from as low as 0.3 SD to as high as 2 SD, indicating a variety of scenarios. In fact, the top 10 generative models out of 2,100 DGPs have error rates within a margin of 0.04%, meaning that it is difficult to distinguish the predictive abilities of these models.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure4, out.width = "100%"}
knitr::include_graphics("../figures/abany02.png")
```
```{=latex}
\end{center}
\caption*{Figure 4: Fake-Data Simulations for Abortion}
\footnotesize{\textit{Notes:} The figure depicts the extent to which specific parameter combinations approximate the observed data. Each point represents our distance measure, calculated as the sum of absolute errors between observed and simulated contingency tables, where darker colors represent smaller distances. We used 5\%, 10\%, and 20\% distance cuts for color shades, along with a second filter using quantiles of 0.05, 0.10, and 0.20 at each level. The shading thus reflects both absolute and relative accuracy among the simulations. The single yellow point represents the parameter combination that best approximates the observed data in the General Social Survey.}
\end{figure}
```

>>>>

---
title: |
  | The Promises and Pitfalls of Using Panel Data to Understand Individual Belief Change[^thanks]
author:
  - Turgut Keskintürk[^TK]
  - Pablo Bello[^PB]
  - Stephen Vaisey[^SV]
date: "`r format(Sys.time(), '%B %Y')`"
output: pdf_document
bibliography: ./helpers/references.bib
csl: ./helpers/asa.csl
abstract: "We investigate whether studies on political belief change can identify change trajectories at the individual level. Using simulations and case studies, we propose a grid-search framework that allows researchers to evaluate the extent to which their target estimates generalize to a study population. We use simulated datasets to estimate plausible values for how many people changed, how much they changed, and who changed, based on observed response trajectories. Our results suggest that researchers should think carefully about the conditions under which they may make claims about belief change at the individual level. We provide an R package to help researchers implement this procedure in their own work."
fontsize: 11pt
fontfamily: mathpazo
geometry: margin=1in
figsintext: yes
header-includes:
- \usepackage{array}
- \usepackage{float}
- \usepackage{caption}
- \usepackage{amsmath}
- \captionsetup{justification=raggedright,
                singlelinecheck=false}
---

@@

This *group-to-person generalizability problem*, the claim that group-level findings may not generalize to each person [@mcmanus2023], has important implications for what we can and can't say about political change across the life-course. While the previous research quantified change with various methods---ranging from assessing the time order of longitudinal observations [@kiley2020; @vaisey2020] to multilevel decomposition of over-time variance [@lersch2023]---these studies have been unable to disentangle whether they find a small amount of change in a large number of people, a large amount of change in a small number of people, or a mix of positive and negative trajectories. In other words, while social science researchers have become sophisticated in evaluating the average levels of change over time in a population, they have mostly overlooked the fact that these average changes can be produced by different processes.

@@

While the cross-sectional processes have received strong scholarly attention in the literature [@vaisey2016; @restrepoochoa2024; @bartels2014], questions of the second sort remained largely unexplored. However, given that theories of political ideology involve theoretical expectations about individual-level outcomes, the fact that we lack information about whether our models capture person-level empirical processes is detrimental to understanding adult political belief change. Our article aims to address this gap by examining whether we can identify (a) *how many people changed*, (b) *how much they changed*, and (c) *who changed* using empirical data.

Using simulations and empirical case studies, we show that it is indeed possible to use panel data to provide plausible estimates of which generative models are more likely to have produced particular empirical observations. We also find, however, that a reliable identification of *who* changed occurs only under very narrow conditions. Hence, the modeling individual change problem is very much like the age-period-cohort identification problem that arises in repeated cross-sectional data [@fosse2019]. In both cases, the problem cannot be "solved" by any particular methodological trick. Thus, our objective is not to provide a methodological panacea to eliminate the problem, but rather to provide a concrete approach that can help us understand the issue more clearly and manage it more effectively.

This approach builds on the "Approximate Bayesian Computation" (ABC) literature [@beaumont2010] and provides a simple grid-search procedure to help researchers evaluate the extent to which their target estimates might be generalized to a study population. Using this procedure, we show that it is possible to approximate to the set of underlying data generation processes for *rate of change*, *strength of change*, and *direction of change* in a panel dataset. We provide an R package, "gridsearch," to help researchers implement our approach in their own empirical applications.

We emphasize the existence of striking heterogeneity in these underlying generative processes even for a seemingly straightforward case like the marijuana legalization in the GSS; that is, we show that theoretically opposite claims---e.g., large swaths of the population increasing their support for the policy, as opposed to a small group of people completely switching their positions---may very well account for the same data. This finding is particularly significant in the longitudinal panel context, which, for most applications, represents the gold standard in evaluating change in observational research. For this reason, we argue that when the meaning of "political change" is not clearly specified, our current empirical strategies could potentially mislead rather than inform, especially if we overlook the underlying generative mechanisms.

We begin with a general overview of the "latent variable model" of beliefs, the problems of reliability and resolution in survey measurement, and the compositional sources of aggregate change. We then introduce our grid-search approach with an illustrating exercise, provide empirical examples, and present a simulation study to explore its implications. Next, we analyze whether this procedure helps us in classification tasks, particularly in evaluating *who* changed. We conclude by discussing the promises and pitfalls of panel data for understanding individual political change.

# Measuring Change in Political Beliefs and Preferences

@@

The latent-variable approach acknowledges a fundamental problem in survey research: one's true score $Y_i$ is almost always realized with a certain degree of measurement error. Call this realization the *observed score*, $y_i$. We often assume that $y_i$ represents the expected value of hypothetical draws from $Y_i$, with an $\epsilon$ term reflecting the measurement error in this process---i.e., $y_i = Y_i + \epsilon_i$. This error might result from various factors that introduce biases in how we extract information from individuals, given that problems like question construction and interviewer effects might influence how individuals reveal their true preferences [@alwin1991]. The inherent randomness in the measurement process implies that each draw from $Y_i$ might at least slightly differ from the true score, and this error term is difficult to circumvent with a one-item measure, which is often the case in large public opinion surveys. Therefore, there is often a degree of systematic gap between $Y_i$ and $y_i$.

@@

This quantity is a natural candidate for describing aggregate belief change, but, under the hood, it summarizes three compositional processes in one metric. Let us define these processes intuitively. The first component is $R$, or the *rate of change*. This represents the proportion of individuals in our dataset with $\delta y_{i, t_0 \to t_1} \neq 0$, i.e., the proportion of individuals who changed in this period. The second component is $S$, the *strength of change*. This represents the "average" magnitude of change among those who changed.[^heterogeneity] The third component is $D$, the *direction of change*. This represents the net directionality of change, i.e., the difference in proportions of positive and negative changers in our observations. If we combine these three components, we can represent $\bar{\delta_y}$ simply as

@@

[^heterogeneity]: Of course, we might further decompose this average strength to a distribution of heterogeneous change trajectories across individuals. For analytical and expository clarity, we bracket this issue in this article.

@@

If we were to believe that *each and every single survey response* represents an accurate mapping of true scores and observed scores, we could seamlessly solve this problem by simply looking at the empirical data. But we contend that the measurement problem does not allow us to do that. The process often fails in at least two ways: (1) our observed measures might have low *reliability*, such that a hypothetical set of repeated draws may significantly differ from one's true score, and (2) the question response categories may be imperfect: the *resolution* of responses may fail to distinguish consequential differences in true scores. Therefore, the average change captured in $\bar{\delta_y}$ may represent a biased or imperfect estimate of change to begin with.

@@

## A Grid-Search Procedure to Adjudicate Data Generation Processes

We build on the Approximate Bayesian Computation (ABC) literature [@beaumont2010] and propose a grid-search procedure to answer this research question. The ABC approach involves simulating parameter values given the data and evaluating the distribution of these parameters that potentially generated our observations. We build an algorithm that

1) Simulates data with several parameters of interest for our DGPs---rate of change, strength of change, direction of change, and reliability of response measurement---using

@@

2) Calculates the distribution of change scores in observed and simulated data, by estimating respondent-level change scores using survey response trajectories at the person level. To put it differently, it first fits, for a binary outcome,

   $$
   \mathbb{P}(y = 1) = \beta_0 + \beta_1t, \ \ t \in [0, 1]
   $$

   where \( t \) is a normalized time variable, and estimates a predicted change score by calculating the difference between the prediction at \( t = 1 \) and the prediction at \( t = 0 \).

3) Compares the distribution of simulated trajectories to the distribution of observed trajectories by calculating a "distance" score---specifically, Kolmogorov-Smirnov distance---to quantify the extent to which specific simulated values reproduce the observed values;

4) Replicates this analysis $N$ times by drawing samples for our main parameters;

5) Inspects the parameter combinations that most closely approximate to the observed values.

@@

@@

[^parameters]: We define additional parameters that we fix in the case studies presented below: $N$ refers to the number of actors in the simulation, $T$ refers to the number of waves we have in the panel study, *balance* refers to the observed marginals in the real datasets, allowing us to "cut" the continuous distribution to 0s and 1s, and *resolution* refers to the number of categories to simplify the latent variable, which is restricted to the binary case in this article.

## Approximating Observed Data via Simulations

Let us flesh out our approach by walking through an illustrative example of how we approximate to the observed empirical responses. Suppose that we *know* the underlying DGP of a dataset of interest. In this dataset, called DGP 1, 500 people are observed over a period of 12 waves. We know that 50% of these respondents make certain changes to their latent positions, and these changes are all in the positive direction with a strength of 1 SD. However, we also know that the item reliability is 80%, which means that, across time, we sometimes miss respondents' true scores. The top-left panel of Figure 2 shows the distribution of change scores from this dataset.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure2, out.width = "100%"}
knitr::include_graphics("../figures/ks_distance.png")
```
```{=latex}
\end{center}
\caption*{Figure 2: Calculating Distance for Observed and Simulated Trajectories}
\footnotesize{\textit{Notes:} The figure presents an illustrative example of two known DGPs: (1) the top-left panel shows the distribution of change scores from DGP 1, (2) the bottom-left panel shows the distribution of change scores from DGP 2, and (3) the right panel shows the empirical cumulative distribution functions from each of these distributions.}
\end{figure}
```

Assume that this dataset is our *observed* dataset. We will now see whether an alternative dataset, called DGP 2, can approximate to its observed parameters. From DGP 1, we know that there are 500 individuals, observed across 12 waves. We also know the marginal distribution of our response variable. We can treat these parameter values fixed for our new simulation. However, we do not know several other parameters: we do not know the *real* rate of change, nor do we know the *real* strength, direction, or reliability. Let us assume, for this specific iteration, a rate of 50% (correct), a strength of 1 SD (correct), a reliability score of 80% (correct), and an equally split direction, half changing positively and half changing negatively (incorrect). The bottom-left panel in Figure 2 shows the distribution of change scores from this particular DGP.

How adequately does DGP 2 represent DGP 1? In the right panel of Figure 2, we show the empirical cumulative distribution functions (ECDFs) for our two distributions, which indicates that DGP 2 does a very poor job of reproducing DGP 1, with a Kolmogorov-Smirnov distance of 0.16. This means that we should be able to improve our predictions significantly if we start tweaking some parameters for DGP 2. Unfortunately, we of course do not know which parameters to tweak; the set of true change parameters is the exact thing we want to derive using observed trajectories. This leads us to consider a wide variety of parameter values that minimize the Kolmogorov-Smirnov distance scores between the observed trajectories and the said simulated trajectories.

In the following sections, we apply this procedure in different contexts. In all cases, we restrict our attention to latent variables with binary outcomes to simplify our analyses. The replication files for the article are stored in https://osf.io/AAAAA. The R package for the use of this approach, "gridsearch," can be accessed via the replication repo or https://anonymized.github.io/gridsearch/.

# Empirical Case Studies

## Exploring the Case of Marijuana Legalization

Let us now return to the marijuana legalization case in the GSS and apply our grid-search procedure to see whether we can detect a set of plausible DGPs for our empirical data.

We first fit 100,000 simulations by drawing samples from four different parameters: *rate of change*, fixed between 0 and 1, *strength of change*, fixed between 0 SD and 2 SD, *balance of change*, fixed between 0 and 1, representing all negative change at 0 and all positive change at 1, and a *reliability score*, fixed between 60% and 100%. We use uniform priors for each parameter. We then calculate the Kolmogorov-Smirnov (KS) distance for each draw and keep the 2% quantile of the DGPs that minimize the KS distance, ending up with 2,000 potential generative models.

Figure 3 shows the distribution of, and bivariate relationships among, these grid-search parameters, suggesting that the posterior distributions of strength and rate parameters are highly variable---and strongly correlated---while the posterior distributions for balance and reliability much more concentrated. Looking at the distributions, the interquartile ranges for strength and rate are 0.63 and 0.42, respectively, while these values are only 0.17 for balance and 0.07 for reliability.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure3, out.width = "100%"}
knitr::include_graphics("../figures/grass02.png")
```
```{=latex}
\end{center}
\caption*{Figure 3: Pair Plots for Marijuana Legalization}
\footnotesize{\textit{Notes:} The figure depicts the distribution of, and bivariate relationships among, grid-search parameter values from the best-fitting DGPs for the case of marijuana legalization (N = 2,000 DGPs from 100,000 simulations).}
\end{figure}
```

These patterns make sense when we examine the conditional relationships among our parameters, presented in Figure 4. We see that the algorithm can easily suggest, at least for this particular case, a specific direction estimate---a median balance of 0.82 with an IQR of 0.73 and 0.90---but it converges to almost analytically determined bands in high-balance panels: we detect either small changes in a large number of people, or large changes in a small number of people. To put this result in perspective, note that the KS distance among these accepted draws are virtually indistinguishable from one another: for instance, the "best" DGP with 40% rate and 0.90 SD strength is within less than 0.001 KS value of a DGP with 20% rate and 1.80 SD strength.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure4, out.width = "100%"}
knitr::include_graphics("../figures/grass03.png")
```
```{=latex}
\end{center}
\caption*{Figure 4: Best-Fitting DGPs for Marijuana Legalization}
\footnotesize{\textit{Notes:} The figure depicts the best fitting 2,000 DGPs (from 100,000 simulations) for the empirical case study. Each point represents an accepted draw. We rounded the balance parameters to show conditional relationships.}
\end{figure}
```

There is one strong implication of this exercise: there are competing generative models that may easily explain the same aggregate patterns, and there is no methodological trick to solve this issue. If we take the measurement error problem to its natural conclusions, the same data can be explained by incompatible theoretical mechanisms, and our conclusions may be wildly off-the-mark.

## Exploring the Case of the Legality of Abortion

What about cases where the direction or even existence of change is less clear-cut? We now turn our attention to the legality of abortion debate, a contentious issue characterized by high disagreement and high political sorting in the American context [@baldassarri2008; @hout2022]. In 2006, 2008, and 2010 GSS panels, respondents agreed or disagreed with the statement: "it should be possible for a pregnant woman to obtain a legal abortion if the woman wants it for any reason." Figure 5 shows response trajectories for this question across three panels, indicating a clear stasis in the marginals.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure5, out.width = "100%"}
knitr::include_graphics("../figures/abany01.png")
```
```{=latex}
\end{center}
\caption*{Figure 5: The Percent Support for the Legality of Abortion}
\footnotesize{\textit{Notes:} The figure depicts the percent of American adults saying "it should be possible for a pregnant woman to obtain a legal abortion if the woman wants it for any reason" in the General Social Survey 2006, 2008, and 2010 panel studies. We included respondents who participated in all three waves of each panel study and weighted the results using GSS post-stratification weights. Note that panel attrition caused compositional changes in the estimates, and compared to the GSS time-series data, the estimates of support are marginally inflated.}
\end{figure}
```

We apply our grid-search algorithm once more, using the same specifications we applied in the case of marijuana legalization: 100,000 draws with $R \in [0, 1]$, $S \in [0, 2]$, and $D \in [0, 1]$, with reliability score ranging between 60% and 100%, each having uniform priors. Once again, we kept 2% of the DGPs. Figure 6 shows the pair plots, while Figure 7 shows the conditional estimates.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure6, out.width = "100%"}
knitr::include_graphics("../figures/abany02.png")
```
```{=latex}
\end{center}
\caption*{Figure 6: Pair Plots for the Legality of Abortion}
\footnotesize{\textit{Notes:} The figure depicts the distribution of, and bivariate relationships among, grid-search parameter values from the best-fitting DGPs for the case of the legality of abortion (N = 2,000 DGPs from 100,000 simulations).}
\end{figure}
```

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure7, out.width = "100%"}
knitr::include_graphics("../figures/abany03.png")
```
```{=latex}
\end{center}
\caption*{Figure 7: Best-Fitting DGPs for the Legality of Abortion}
\footnotesize{\textit{Notes:} The figure depicts the best fitting 2,000 DGPs (from 100,000 simulations) for the empirical case study. Each point represents an accepted draw. We rounded the balance parameters to show conditional relationships.}
\end{figure}
```

We find strong variability in the plausible generative scenarios. Note how, in all cases, the accepted draws are squished towards 0---either via 0 rate of change or via 0 strength of change---indicating a clear preference towards no change in the underlying processes. That said, the plausibility bounds also become strongly noisy when the estimated balance is even: accepting scenarios ranging from no change to the now-familiar high change in a low number of people and low change in a high number of people. The highest KS distance scores among all the accepted samples is less than 0.01, suggesting that all draws fit reasonably similarly across all scenarios.

This exercise is an appropriate illustration of the limitations of panel studies. Based on the observed data, we cannot determine whether the observed marginals result from a small number of people making substantial belief changes or a large number of people making minor political changes. This may, ultimately, boil down to the issue of observation frequency: with three-wave panel data and reasonable doubts about the accuracy of every single survey response due to reliability and resolution concerns, it is challenging to adjudicate the underlying generative models. In the next section, we relax this assumption and extend our method to cases with more than three waves.

## Exploring the Case of Immigration Spending

The data for this exercise comes from the *Political Psychology Data from a 26-Wave Yearlong Longitudinal Study* [@brandt2021], conducted biweekly in 2019-2020 for a year. In one question in the survey, respondents were asked whether they agree with the policy of increased federal spending on immigration control. We binarized the responses to this question such that those agreeing with the statement received a 1, and those who said the spending should be decreased or kept the same received a 0. We then dropped respondents with missing data and trimmed the periods to pre-COVID windows, resulting in 213 individuals observed 17 times over 34 weeks. Although the sample for this case study is obviously not representative of any national population, it provides an intensive set of repeated observations at the individual level. Figure 8 shows the trajectory of average responses across these 17 waves within the study window.

```{=latex}
\begin{figure}[!ht]
\begin{center}
```
```{r figure8, out.width = "100%"}
knitr::include_graphics("../figures/immigration01.png")
```
```{=latex}
\end{center}
\caption*{Figure 8: The Percent Support for an Increase in the Control of Immigration Spending}
\footnotesize{\textit{Notes:} The figure depicts the percent support for the statement "the federal spending to control immigration should be increased" in the Political Psychology Data from a 26-Wave Yearlong Longitudinal Study.}
\end{figure}
```
